Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-05-27 paper/main.tex
changes through xi

\IEEEPARstart{V}isualization software libraries consist of a set of functions for translating data into visual representations. These functions are expected to preserve the structure of the input data, however that structure is specified. Over time, library developers may need to develop new components to adapt to changing data needs, especially the growth of complex structured and distributed data. We propose that the complex structure of data can be described using concepts from algebraic topology and category theory, and that category theory can be used to express the constraints these components must satisfy in order to produce structure preserving data to visual transformations. This abstraction of the data structure and specification of the components can then be used to guide the developement of functional\footnote{functional programming paradigm} components. Since these components are functional, they are expected to be more composable and modular than existing components, have fewer side effects and therefore be easier to incorporate alignside existing library elements, and be verifiable and therefore easier to test \cite{huHowFunctionalProgramming2015,hughesWhyFunctionalProgramming1989}. In this paper, we introduce a mathematical framework for describing, constructing, and testing visualization library components and provide a toy example illustrating these concepts.

\section{Related Work}
Visualization library components are required to preserve structure; in a visualization context this has widely come to be defined as a combination of the topological properties of the dataset and the mathematical structure of the dataset fields.

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. For example, a 1D continuous function is evaluated on a 1D continuous interval, and a map is 2D continuous surface. A network is a union of discrete nodes and continuous lines. The nodes represent the continuity of a table of discrete weather station data and the graph edges are encoded with the distances between the nodes. \label{fig:related-work:continuity:ktypes}}
\end{figure}
The topological properties of a dataset are the ways in which elements are organized, as described by Wilkenson\cite{wilkinsonGrammarGraphics2005}. Some examples, as shown in \autoref{fig:related-work:continuity:ktypes} are discrete rows of a table, a continuous 1D line, or a surface. The topological space (in purple) and the data form a key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. For example  in \autoref{fig:related-work:continuity:ktypes}, a point on the green line is mapped to by evaluating the gaussian on a point on the interval and each pixel in the map can be found via a lookup on the 2D surface. This general formulation of topology gives us a language for expressing the continuity of a dataset. The network in \autoref{fig:related-work:continuity:ktypes} is a union of a 0D discrete points and 1D lines connecting those points. The data record in the yellow row can be expressed as a constant function of the yellow network node, while the data on the graph edge between any two points is a function of the distance between those points.

\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{continuity.png}
  \caption{The \texttt{line} and \texttt{image} visual algorithms take in the same array of data. The \texttt{line} assumes that the array is a list of lines, while the \texttt{image} assumes that the array is an image.  \label{fig:related-work:continuity:borked}}
\end{figure}
The continuity matters to visualization libraries because visual algorithms, such as a the line plot and image plot in \autoref{fig:related-work:continuity:borked}, assume the continuity of their data, as described in taxonomies of visualization algorithms by Troy, M\"{o}ller, and Chi \cite{toryRethinkingVisualizationHighlevel2004,chiTaxonomyVisualizationTechniques2000}. In \autoref{fig:related-work:continuity:borked}, both \texttt{line} and \textt{image} algorithms are fed the same input matrix. The line plot assumes this matrix is a list of lines and plots it as such. The image map assumes that the data is an image and plots it as such. Neither function has a way of asking the matrix what it is, nor does the matrix encode information about the continuity of the data it contains. While this work is not about choosing the right visualization type for the given continuity, the ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of ensuring that the assumptions match.

Current visualization libraries handle continuity in two main ways. Libraries that provide "building block" \cite{wongsuphasawatNavigatingWideWorld2021} components for building new visual algorithms embed their assumptions in their algorithms. This flexibility allows libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007}, VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012}, and D3\cite{bostockDataDrivenDocuments2011} to support a wide variety of data types, but at the cost of the ambiguity of \autoref{fig:related-work:continuity:borked}. These libraries can also have visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. On the other hand, domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}. Tabular data is a very common data topology and is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

\note{maybe move the following to \autoref{sec:atct:io} and consolidate?}
To bridge the need for flexibility and cohesiveness, we build on the proposal of fiber bundles as a general abstraction model for visualization data, as proposed by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}. Fiber bundles are a mathematical structure that can be used to encode the topological properties of visualization data. There is also a methodology for encoding field types in fiber bundles, as described in Spivak's description of tables as fiber bundles\cite{spivakSIMPLICIALDATABASES}. Fiber bundles are discussed in more detail in \autoref{sec:atct:fiber-bundles}, but broadly they provide a mathematical representation of data that can encode topological properties, field types, and data values in a way that is both uniform and continuity and dimension independent way.

\subsection{Equivariance}
Equivariant maps are functions where variations in the input match variations in the output. The expectation that visualizations are equivariant maps from data to visual representation was first formalized by Bertin \cite{bertinSemiologyGraphicsDiagrams2011a}. He classified datasets into point, line, and surface topologies and proposed that these datasets should be represented by visual elements with the same sort of topology - points should be represented by discrete entities, lines by lines, and surfaces through area marks. Bertin also codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}.

The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that changing the data should change the visualization in the same way in a perceptually significant manner. While the data representation introduced in \autoref{sec:atct:io} dispenses with a distinction between containers and the data, the definition of equivariance introduced in  \autoref{sec:artist:equivariance} includes invariant transformations and visually measurable transformations.

Visualization libraries are in part measured by how expressive the components of the library are; expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving automated tool, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Some libraries, such as Matplotlib, set default encoders for their visual algorithms and allow the user to change the default. In this work, we propose guidelines for building expressive structure informed data to visual transformers. As with visual algorithms, we aim to provide a way to check that the structure of the data matches the structure expected by the component. Doing so allows users to more easily develop equivariant visual algorithms.

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
A goal of this work is to develop guidelines for building composable structure preserving functions for translating data into visual representations. Concepts in algebraic topology and category theory provide a language for formally expressing the structure that these visualizations are expected to preserve. Fiber bundles and sheaves, which come from algebraic topolology, provide a uniform generalizable way to describe data and visualization. Category theory is a method to describe how objects are specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}\; therefore in this work we propose categorical constructions of visualization components as specifications for implementable components. A brief introduction to category theory for visualization practitioners is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
Fiber bundles, are a "unified, dimension-independent framework", as described by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, that allow us to seperatly describe the topological properties and fields of a data source and also the mapping between topolology and data values. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepointc \in \dbasec$ and topology $\mathcal{T}_{\dbasec}$; $\mathcal{T}_k$ is a set of opensets $ \dbasepointc \in \opensetc \subseteqq \dbasec$ \cite{munkresElementsAlgebraicTopology1984} that cover the base space \dbasec.  The \textcolor{fiber}{fiber space} is a topological space \dfiberc\ that is the preimage of the projection function $\pi$ at a point in the base space $\dfiber_{\dbasepointc} = \pi^{-1}(\dbasepointc)$ and fibers of a bundle are isomorphic $\dfiberc \simeq \dfiberc_{\dbasepointc}\;\forall \dbasepointc \in \dbase$. The \textcolor{section}{sections} of a bundle $\dsection$ are maps from the base space to points in the bundles over that base space $\textcolor{set}{\Gamma}$ denotes the set of all sections in a bundle over an openset. 
\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}

@@

The bundle in \autoref{fig:atct:stripbundle} is non-trivial, meaning that the fiber is not identical over every point in the base space $\dbase$. One section over this 1D continuous space is a sin function that takes as input a point in the base space $\dbase$ and returns the sin of the point.

A fiber bundle is locally trivial, which means that for every point \dbasepointc\ there exists an open neighborhood $\dbasepointc \in \opensetc \subseteq \dbasec$ such that there is a homeomorphism $\pi^{-1}(\opensetc)\xrightarrow{\equivc} \opensetc \times \dfiberc$. A bundle may be globally trivial, meaning that $\dtotalc = \dbasec \times \dfiberc$, and generally can be thought of as a twisted product of total and base spaces \cite{munkresElementsAlgebraicTopology1984}. Throughout this work we will assume that fiber bundles are trivial because that allows us to model the section functions as from $\texttt{base} \rightarrow \texttt{fiber}$ since the fiber is the same at all points in the base space.

Fiber bundles can be used to model data for visualization, as proposed by Butler\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We describe the topological properties of the data \cite{wilkinsonGrammarGraphics2005} as the base space \dbasec. The topological properties, as described by Wilkenson, are how the data is organized - for example \dbase\ could be discrete points, a line, a 2D surface, a 3D volume, or a graph. The names and data types of the data fields can be expressed as the fiber space \dfiberc, as described by Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}. Extending Spivak's fiber bundle description of a table, we define the values returned by a section as named and typed records of the dataset.

@@

The bundle in \autoref{fig:atct:networkbundle} is constructed from a union of bundles over discrete points $\dbase_{1}, \dbase_{3}$ and a 1D continuous connecting line $\dbase_{2}$ that connects the two points.

A fiber bundle is a mathematical structure for encoding continuity, type, and datasets with that continuity and type. In \autoref{fig:atct:fb}, the continuity is a discrete point \dbase. The fiber space that encodes the types is the cartesian product of the field types, and the section returns a single record with values of each field type. While this is a very trivial example, a fiber bundle base space can have almost any topology. The fiber space also does not really have restrictions on which types it can encode or the structure of those types - for example, \texttt{speed} and \texttt{direction} are separate fibers in \autoref{fig:atct:fb} but can be combined into a single two dimensional fiber $\dfiber_{speed, direction} = {\mathbb{R}^{+}}^{s}$ The generality in base space and fiber space makes fiber bundles a unified dimension and type independent data abstraction that can encode most of the common visualization data set structures, such as continuous functions, tables, data cubes, images and networks.

@@

\subsection{Topology and Field Categories}
\label{sec:atct:io}
We define categorical types for the topology and the fibers because categories are defined both by the objects of the category and the valid morphisms on the object. This allows for the formal construction of an equivariance aware type that holds the topology, field, type, and what changes to either must be preserved.

@@

\subsection{Formal Properties of Data Containers}
\label{sec:atct:sheaves}
Sheaves on bundles are an "algebraic data structure", as described by Ghrist \cite{ghristElementaryAppliedTopology2014}, for expressing the bookkeeping involved in keeping track of sections of data over opensets. Sheaves are a way of expressing that subsets, distributed data, and streaming data are local sections that can be glued together. Formally, pre-sheaves are functors, meaning they are functions from objects of one category to objects of another category \cite{WhatFunctorDefinitions}. Here the sheaf 
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
denotes a sheaf from opensets on the base space $\dbase$ to sets of sections on the bundle $\dtotal$. As a contravariant functor it preserves morphism s

\begin{equation}
  \label{eq:atct:presheaf}
  \begin{tikzcd}
    \cgamma{\opensetc_1}{\dtotalc\restriction_{\opensetc_1}}  &  & \cgamma{\opensetc_2}{\dtotalc\restriction_{\opensetc_2}} 
    \arrow[ll, "\iota^*"', maps to, color=set] \\
    & & \\
    \opensetc_1 
    \arrow[rr, "\iota", maps to, color=base] 
    \arrow[uu, "{\sheafc_{\dbasec, \dtotalc}}", maps to, color=sheaf] &  & \opensetc_2 
    \arrow[uu, "{\sheafc_{\dbasec, \dtotalc}}"', maps to, color=sheaf]              
    \end{tikzcd}
\end{equation}

which associates a set of sections $\cgamma{\openset}{\dtotal\restriction_{\openset}}$ with an openset \openset. The sets of sections are objects of the category \setc\ and the morphisms of the category are restrictions\footnote{restriction functions } $\iota^*$. The sheaf over a an open set $\openset$ surrounding a point $\dbasepoint$ is called a stalk\cite{StalkSheaf2019}
\begin{equation}
  \label{eq:atct:sheaf:stalk}
    \sheaf_{\dbase, \dtotalc}\restriction_{\dbasepoint}\coloneqq \lim\limits_{\openset\ni \dbasepoint} \Gamma(\openset, \dtotal\restriction_{\openset}) 
\end{equation}
where the fiber is contained inside the stalk  $\dfiber_{\dbasepoint} \subset  \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. The germ is the section evaluated at a point in the stalk  $\dsection(\dbasepoint) \in \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$ and is the data. \note{I never use stalk and germ terminology at any point in this paper}

>>>>

\IEEEPARstart{V}isualizations are expected to be a faithful translation of the inherent structure of the data being visualized; therefore the components of visualization software libraries that translate data to graphics are also expected to preserve this structure. Over time, library developers may need to develop new components to adapt to changing data needs, especially high dimensional multivariate and distributed data. 
This paper provides a method for formally describing structure that should be preserved and this work provides guidance on how to verify this preservation. This abstraction of the data structure and specification of the components can then be used to guide the developement of functional\footnote{functional programming paradigm} components. Since these components are functional, they are expected to be more composable and modular than existing components, have fewer side effects and therefore be easier to incorporate alignside existing library elements, and be verifiable and therefore easier to test \cite{huHowFunctionalProgramming2015,hughesWhyFunctionalProgramming1989}. In this paper, we introduce a mathematical framework for describing, constructing, and testing visualization library components and provide a toy example illustrating these concepts.

\section{Related Work}
We present an abstraction of data that is agnostic to the types of fields or number of dimensions of the data by building on work that has codified structure and how it is preserved. Structure is mainly described as a combination of the topological properties of the data and the mathematical structure of the dataset fields. Structure preservation is expressed as equivalent changes on the data input and graphics output of a visualization algorithm. Our model encapsulates all these notions of structure by providing a language for expressing topology, field structure, and changes to the data model.

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. For example, a 1D continuous function is evaluated on a 1D continuous interval, and a map is 2D continuous surface. A network is a union of discrete nodes and continuous lines. The nodes represent the continuity of a table of discrete weather station data and the graph edges are encoded with the distances between the nodes. \label{fig:related-work:continuity:ktypes}}
\end{figure}
The topological properties of a dataset are the ways in which elements are connected, as described by Wilkenson\cite{wilkinsonGrammarGraphics2005}. Some examples, as shown in \autoref{fig:related-work:continuity:ktypes} are discrete rows of a table, a continuous 1D line, or a surface. The topological space, in purple, acts as a generalized indexing space for the data. For example, a point on the green line is mapped to by evaluating the gaussian on a point on the interval and each pixel in the map can be found via a lookup on the 2D surface. This general formulation of topology gives us a language for expressing the continuity of a dataset. The network is a union of a 0D discrete points and 1D lines connecting those points. The data record in the yellow row can be expressed as a constant function of the yellow network node, while the data on the graph edge between any two points is a function of the distance between those points.

Bertin stated that data should be represented by visual elements with the same type of topology - points should be represented by discrete entities, lines by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011a}.  Visual algorithms assume the continuity of their data, as described in taxonomies of visualization algorithms by Troy, M\"{o}ller, and Chi \cite{toryRethinkingVisualizationHighlevel2004,chiTaxonomyVisualizationTechniques2000} and try to preserve it, but many data structures do not provide a method for encoding the topology of the data they store. For example, a \texttt{line} algorithm often take as input a list of (x,y) coordinates. The \texttt{line} algorithm assumes that its input is continuous and therefore displays the data as if it was continuous, regardless of the actual data topology. The ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of verifying that the assumptions match and therefore that the components are preserving the topological structure.

\subsection{Equivariance}
Visualizations that preserve the mathematical structure of the data are called equivariant maps. Equivariant maps are maps where every change to an input has an equivalent change in the output. Bertin codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element \cite{bertinSemiologyGraphicsDiagrams2011a}. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}. The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should coresspond to changes in the visualization in a perceptually significant manner. While the data representation introduced in \autoref{sec:atct:io} dispenses with a distinction between containers and the data, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi} and the definition of equivariance introduced in \autoref{sec:artist:equivariance} includes the correspondence between data and visual changes.

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are; expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
In this section, we propose a mathematical abstraction of a data container and graphic prerendered output. This mathematical abstraction has a robust language for expressing topology and fields; expresses how to verify that data continuity is preserved on subset, distributed, and streaming data; and formalizes the expectation of a correspondence between data and visual elements.

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
To bridge the need for flexibility and cohesiveness, we model data using a mathematical representation of data that can encode topological properties, field types, and data values in a way that is both uniform and continuity and dimension independent way. This mathematical representation is called a fiber bundle and its use as an abstract data representation in visualization was first proposed by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$; $\mathcal{T}_\dbasepoint$ is a set of opensets $ \dbasepoint \in \openset \subseteqq \dbase$ \cite{munkresElementsAlgebraicTopology1984} that cover the base space \dbasec.

The \textcolor{fiber}{fiber space} is a topological space \dfiberc\ that is the preimage of the projection function $\pi$ at a point in the base space $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$, meaning that all points in a fiber over a space $\dfiber_{\dbasepoint}$ map to the same point $\dbasepoint$. Fibers of a bundle are isomorphic $\dfiberc \simeq \dfiberc_{\dbasepointc}\;\forall \dbasepointc \in \dbase$, which means that the space of possible values is the same set of values over all points in the base space.  The fiber space can be used to encode the space of all possible values of a data field, which is analogous to a CS data type, and the fibers can be named, as described by Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.

The \textcolor{section}{sections} $\dsection$ of a bundle are continuous functions that bind the base space with a set of values in the total space $\dtotal$ and $\textcolor{set}{\Gamma}$ denotes the set of all sections in a bundle over an openset.

\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
Each section function takes as input a point in the basespace $\dbasepoint$ and returns a single point in the fiber. Extending Spivak's fiber bundle description of a table, we define the values returned by a section as named and typed records. Since the fiber space is a topological space, it does not really have restrictions on the dimension of the types it can encode or the structure of those types. For example color can be represented as the cartesian product of three distinct sets, one for each color, or a three dimensional topological space $\mathbb{R} \times \mathbb{R} \times \mathbb{R} = \times \mathbb{R}^{3}$.

The base space and the section evaluated on the base space form a sort of key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. Therefore, we do not impose any semantic constraints on what the fiber and base space encode. For example, a data cube of temperature can be encoded such that every fiber is a set of 2D temperature surfaces $\dfiber_{tempsurf}$, the base space is a 1D continuous interval, and each section returns a surface $\dsection(\dbasepoint) = f(x,y)$ where x and y are already parameterized. Alternatively, the fiber can encode each cell of the data cube $\dfiber_{lat} \times \dfiber_{lon} \times \dfiber_{temp}$, the base space could be a 3D volume, and each section returns a $\dsection(\dbasepoint) = \{lat, lon, temp\}$ record.

Fiber bundles where the fibers are not identical are called non-trivial bundles. Non-trivial bundles allow us to generalize bundles to many different types of total spaces. We can also decompose non-trivial bundles into trivial bundles, which are bundles where all the fibers are identical. This is because all fiber bundles are locally trivial, which means that for a small enough subset of the data $\opensetc \subseteq \dbasec$, all the fibers are equivalent $\pi^{-1}(\opensetc)\xrightarrow{\equivc} \opensetc \times \dfiberc$. A bundle may be globally trivial, meaning that $\dtotalc = \dbasec \times \dfiberc$, and generally can be thought of as a twisted product of total and base spaces \cite{munkresElementsAlgebraicTopology1984}. Using trivial and locally trivial bundles allows us to encode the possible data records in a bundle as a single fiber $\dfiber$ type rather than having to maintain a fiber type for each point \dbasepoint.

@@

The bundle in \autoref{fig:atct:stripbundle} is a fiber bundle representation of a mobius band, which is the total space \dtotal\ of the bundle. Each fiber \dfiber\ collapses to a point in the base space via the projection function $\pi$. The bundle is non-trivial because the fibers are not identical $[-1, 1] \not \equiv [1, 1]$. One section over this 1D continuous space is a \texttt{sin} function that takes as input a base point $\dbasepoint_{i}$ and returns a point in the fiber $sin(\dbasepoint_{i})$.

@@

In \autoref{fig:atct:networkbundle}, the network from \autoref{fig:related-work:continuity:ktypes} is encoded as a section of a bundle. This bundle is constructed from a union of bundles over discrete points $\dbase_{0}, \dbase_{2}$ and a 1D continuous connecting line $\dbase_{1}$ that connects these two points. The fiber over the data points $\dfiber_{node}$ expresses that the name column type is a subset of \textttt{string}, the temperature column type is \texttt{float}, and the precipitation column type is positive \texttt{float}. The fiber over the network edge encodes that the distance from the edge table has type \texttt{float}. The data section evaluated at a discrete point $\dbase_{0}$ returns the values for the first row in the table. Evaluating the function on $\dbase_{2}$ would return a different row in the table. The section $\dsection$ knows which nodes $\dbase_{1}$ is between and so returns the corresponding data value.  The network edge function is constant and returns the same value any where along $\dbase_{1}$.

The only constraint on sections is that they are (piecewise) continuous and that they return values in the fiber. They can be constant or non-constant and they can be defined over a single type of topology or a union of topologies. The generality in base space, fiber space, and section makes fiber bundles a unified dimension and type independent data abstraction. Therefore fiber bundles provide guidance on constructing a structure aware data interface layer that can support most of the common visualization data set structures, such as continuous functions, tables, data cubes, images and networks.

@@

\subsection{Topology and Field Categories}
\label{sec:atct:io}
We define categorical types for the topology and the fiber because this allows us to describe  how objects that encode topology and fiber should be specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}, which we can then translate to the implementation of these objects. In particular, the categorical construction allows us to bundle the topology and fiber as objects together with the changes to these objects, which are termed morphisms, that the visualization is expected to preserve.

@@

A brief introduction to category theory for visualization practitioners is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

\subsection{Formal Properties of Data Containers}
\label{sec:atct:sheaves}
Sheaves are a mathematical way to describe the expectation that data structures implement the bookkeeping that keeps track of the continuity of the data and graphic \cite{ghristElementaryAppliedTopology2014}. These constraints generalize to data that is subsetted, distributed, streaming, and on-demand.

Presheaves are functors, meaning they are functions from objects of one category to objects of another category\cite{WhatFunctorDefinitions} that preserve the morphisms in each category. By definition, presheaves are maps from objects in an arbitrary category to an object in the category \setb\cite{nlab:presheaf}. Here the arbitrary category is the base space category and the object in \setb is the set of sections

\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}

Presheaves are a contraviant functor, which means that the inclusion morphisms $\iota$ between base space objects go in the opposite direction from the restriction morphisms $\iota^*$ between objects in set.  
\begin{equation}
  \label{eq:atct:presheaf}
  \begin{tikzcd}
    \cgamma{\opensetc_1}{\dtotalc\restriction_{\opensetc_1}}  &  & \cgamma{\opensetc_2}{\dtotalc\restriction_{\opensetc_2}} 
    \arrow[ll, "\iota^*"', maps to, color=set] \\
    & & \\
    \opensetc_1 
    \arrow[rr, "\iota", maps to, color=base] 
    \arrow[uu, "{\sheafc_{\dbasec, \dtotalc}}", maps to, color=sheaf] &  & \opensetc_2 
    \arrow[uu, "{\sheafc_{\dbasec, \dtotalc}}"', maps to, color=sheaf]              
    \end{tikzcd}
\end{equation}
A presheaf describes the expectation that any set of sections defined over a larger space $\dsection \in Gamma(\opensetc_2, \dtotal\restriction_{\openset_{2}})$ is also defined over subspaces of that space $\dsection\restriction_{\opensetc_{1}}$.

The locality and gluing axioms \cite{bakerMathsSheaf} of a sheaf describe the conditions for verifying that the sections over subspaces are part of a cohesive section over the whole space. The locality axiom asserts that two sections in a sheaf are equal when they evaluate to the same values over the same set of open sets. The gluing axiom asserts that the union of sections defined on specific open sets is equivalent to one big section over the union of spaces. The gluing axiom therefore expresses an equivalency between a concatenated representation of data and the distributed representation of the same data. In \autoref{fig:atct:networkbundle} this axiom means that the sections over each node and the edge can be combined into one section over the piece of the network in the figure.

Each section of a sheaf over a point returns a single record in the fiber. The sheaf over an open set $\openset$ surrounding a point $\dbasepoint$ is called a stalk\cite{StalkSheaf2019}
\begin{equation}
  \label{eq:atct:sheaf:stalk}
    \sheaf_{\dbase, \dtotalc}\restriction_{\dbasepoint}\coloneqq \lim\limits_{\openset\ni \dbasepoint} \Gamma(\openset, \dtotal\restriction_{\openset}) 
\end{equation}
where the fiber is contained inside the stalk  $\dfiber_{\dbasepoint} \subset  \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. The germ is the section evaluated at a point in the stalk  $\dsection(\dbasepoint) \in \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$ and is the data. The stalk and germ in the stalk provide the derivative of the record, which is needed for some computational and visualization tasks.

\subsection{Mathematical Structure of Data}
\label{sec:atct:sheaves:measurments}
We define structure on data as the allowable \textcolor{action}{transformations on the bundle}, which we formalize as actions $\dfuncc$ on sections of that sheaf. Actions  $act: X \times Y \tightarrow Y$ are elements of one set $X$ applied to elements of another set $Y$ such that the action in X applied to Y returns an element in Y\cite{nlab:action}. Actions are a generalization of applying a function to an element and returning an element of the same type. Our sheaf action  $\dfunc$ is a generalization of the classification of measurement scales by the actions allowed on the data values\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}, which often serves as the basis of the equivariance described in \autoref{sec:related-work:equivariance}.

We separate data transformations into two types, transformations on the base space $(\dfunchc, \dfuncpullc)$ and transformations on the fiber space $\dfunctc$. These diagrams are seperated because the base space transformation and the pullback are functors on the openset object and associated set of sections, while the fiber transformation takes a single fiber to a different fiber in the set of sections.
